---
output: github_document
always_allow_html: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# biodiscvr: Biomarker Discovery Using Composite Value Ratios

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
[![R-CMD-check](https://github.com/isaac-6/biodiscvr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/isaac-6/biodiscvr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->


## Overview

`biodiscvr` provides a framework for discovering and evaluating novel or optimized biomarkers defined as ratios of composite values derived from feature sets (e.g., regional measurements from imaging data). It is particularly suited for analyzing longitudinal multi-cohort datasets.

The core functionality utilizes a Genetic Algorithm (GA) to search the feature space for optimal numerator and denominator combinations based on biomarker performance metrics calculated using linear mixed-effects models (Repeatability, Group Separation, Sample Size Estimates).

## Installation

You can install the development version of `biodiscvr` from [GitHub](https://github.com/isaac-6/biodiscvr) with:

```r
if (!require("remotes")) install.packages("remotes")
remotes::install_github("isaac-6/biodiscvr", build_vignettes = TRUE)
```

**Note:** You may need Rtools (Windows) or Xcode Command Line Tools (macOS) installed. The installation will also attempt to install required packages listed under `Imports` (like `yaml`, `readr`, `dplyr`, `lme4`, `longpower`, `GA`, `rlang`).

## Getting Started: Workflow Tutorial

The best way to learn how to use the package is through the included vignette, which walks through the entire workflow step-by-step using your own data alongside the package's default configuration files.

After installing the package, you can access the vignette with:

```{r view-vignette, eval=FALSE}
# Load the package first if needed
# library(biodiscvr)

# View the main workflow vignette
vignette("biodiscvr", package = "biodiscvr")
```

The vignette covers:
*   Setting up your data directories and output paths.
*   Loading data with `load_datasets()`.
*   Checking and preparing data with `check_and_prepare_data()`.
*   Preprocessing data with `preprocess_datasets()`.
*   Generating demographic tables with `create_demographics_table()`.
*   Running discovery experiments with `run_experiments()`.
*   Evaluating literature biomarkers with `evaluate_literature_biomarkers()`.

## Core Functions

The package provides several key functions for the workflow:

*   `load_datasets()`: Loads data from structured directories.
*   `check_and_prepare_data()`: Validates data, prepares SUV columns based on dictionary matching/renaming.
*   `preprocess_datasets()`: Filters data based on minimum entries and inclusion criteria.
*   `create_demographics_table()`: Generates demographic summary tables.
*   `run_experiments()`: Orchestrates single- and multi-cohort discovery runs based on an experiments configuration file.
*   `evaluate_literature_biomarkers()`: Evaluates predefined biomarkers on the processed data.
*   `biodiscvr_single()`: Runs the CVR discovery process for a single dataset (typically called by `run_experiments`).
*   `biodiscvr_multicohort()`: Runs the CVR discovery process optimized across multiple datasets (typically called by `run_experiments`).

Consult the help page for each function (e.g., `?run_experiments`) for detailed argument descriptions.

## References

This package builds upon the methodologies described in:
- Llorente-Saguer, I. et al. (2024). *A data-driven framework for biomarker discovery applied to optimizing modern clinical and preclinical trials on Alzheimer’s disease*. Brain Communications, Volume 6, Issue 6. [doi link](https://doi.org/10.1093/braincomms/fcae438)

## Citation

If you use `biodiscvr` in your research, please cite it. You can get the citation information by running: `citation("biodiscvr")` after installation.

## Contributing

Please note that the 'biodiscvr' project is released with a Contributor Code of Conduct (`code_of_conduct.md`). By contributing to this project, you agree to abide by its terms.

## License

This package is licensed under the GPL v3 License. See the [LICENSE](LICENSE.md) file for details.